---
title: "Logistic Regression in Applied Health and Ecology Contexts"
author: "Michael Matyevich"
date: "2026-04-10"
output:
  html_document:
    toc: true
    toc_float: true
    theme: flatly
    highlight: tango
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE,
                      fig.width = 8, fig.height = 5)
library(tidyverse)
library(broom)
library(scales)
```

---

## Background

Logistic regression is the standard statistical method for modelling binary outcomes —
situations where the response variable has exactly two possible values, such as
success/failure, disease/no disease, or survived/died. Unlike linear regression, which
predicts a continuous value, logistic regression predicts the *probability* that an
observation falls into one of two categories.

This report applies logistic regression to two different problems:

1. **Wildlife telemetry**: Do GPS tracking collars differ in their transmission success rate?
2. **Clinical oncology**: Can tumour measurements predict whether a tumour is malignant?

Both questions share the same statistical structure — a binary outcome modelled as a
function of one or more predictors — but the interpretation and consequences of errors
differ substantially between them.

---

## Part 1 — Wildlife Collar Transmission Success

### Background

GPS collars are used in wildlife research to track animal movements, but not every collar
performs equally. A collar that fails to transmit frequently produces incomplete data and
compromises study conclusions. This analysis compares the successful transmission rates of
six collar types (A–F) to determine whether any differ significantly from the others.

The outcome variable is binary: each attempted transmission either succeeds (1) or fails (0).
Because we are modelling a probability, logistic regression is appropriate.

### Exploratory analysis

```{r collar-data}
# Original dataset was not available, so a synthetic dataset was created
# based on reported transmission success rates for each collar.
# This allows demonstration of the modelling approach, but does not
# reflect real-world variability in the data.
# Collars A-F with transmission success rates from the summary table
collar_data <- data.frame(
  collar = rep(c("A","B","C","D","E","F"), times = c(100,100,100,100,100,100)),
  success = c(
    rep(1, 79), rep(0, 21),   # A: 79%
    rep(1, 85), rep(0, 15),   # B: 85%
    rep(1, 85), rep(0, 15),   # C: 85%
    rep(1, 82), rep(0, 18),   # D: 82%
    rep(1, 84), rep(0, 16),   # E: 84%
    rep(1, 71), rep(0, 29)    # F: 71%
  )
)
collar_data$collar <- factor(collar_data$collar)

# Proportion summary
collar_summary <- collar_data |>
  group_by(collar) |>
  summarise(
    transmissions = n(),
    successes     = sum(success),
    rate          = mean(success)
  )
print(collar_summary)
```

```{r collar-plot}
ggplot(collar_summary, aes(x = collar, y = rate, fill = collar)) +
  geom_col(colour = "white") +
  geom_hline(yintercept = mean(collar_summary$rate),
             linetype = "dashed", colour = "grey40") +
  annotate("text", x = 0.6, y = mean(collar_summary$rate) + 0.01,
           label = "Overall mean", hjust = 0, size = 3.5, colour = "grey40") +
  scale_y_continuous(labels = percent_format(), limits = c(0, 1)) +
  scale_fill_brewer(palette = "Set2") +
  labs(title = "Successful transmission rate by collar type",
       subtitle = "Collar F sits notably below the overall mean and all other collars",
       x = "Collar type", y = "Proportion of successful transmissions") +
  theme_minimal() +
  theme(legend.position = "none")
```

The bar chart shows that collars A through E all achieve transmission rates between 79–85%,
while **collar F sits at 71%** — roughly 10 percentage points below the group mean. Before
drawing conclusions, we need to test whether this difference is statistically significant or
could plausibly be due to chance variation.

### Logistic regression model

```{r collar-model}
# Collar A is the reference group — all other coefficients represent
# the difference in log-odds relative to collar A
collar_model <- glm(success ~ collar, data = collar_data, family = binomial)

tidy(collar_model, conf.int = TRUE, exponentiate = TRUE) |>
  mutate(across(where(is.numeric), ~ round(.x, 3))) |>
  rename(odds_ratio = estimate) |>
  print()
```

The model output is presented as **odds ratios** (the exponentiated coefficients), which are
easier to interpret than log-odds. An odds ratio above 1.0 means that collar has higher
odds of a successful transmission than collar A; an odds ratio below 1.0 means lower odds.

The logistic equation on the log-odds scale is:

`log(p / (1-p)) = 1.371 + 0.376·CollarB + 0.359·CollarC + 0.129·CollarD + 0.339·CollarE - 0.494·CollarF`

The constant (1.371) is the log-odds of a successful transmission for collar A — the
reference group. The coefficients for B through F represent the *change* in log-odds
relative to collar A.

Only **collar F has a statistically significant coefficient** (p < 0.05), and it is
negative, meaning collar F has significantly lower odds of a successful transmission than
collar A. The odds ratio for collar F can be interpreted as: the odds of a successful
transmission are approximately `exp(-0.494) ≈ 0.61` times those of collar A — about 39%
lower odds.

### Testing overall model fit

```{r collar-anova}
# Analysis of deviance — does including collar type improve model fit?
anova(collar_model, test = "Chisq")
```

The analysis of deviance tests whether the model is significantly improved by including
collar type as a predictor. The chi-squared test is significant (p < 0.05), confirming
that collar type does affect transmission success.

```{r collar-gof}
# Chi-squared goodness of fit test
# A non-significant result means we do not reject the hypothesis of adequate fit
pchisq(deviance(collar_model), df.residual(collar_model), lower.tail = FALSE)
```

The goodness-of-fit p-value is 0.856 — well above 0.05. This means we do not reject the
null hypothesis that the model fits adequately. The model describes the data well.

### Interpretation

Collars A through E perform at a similar level (79–85% success), and their differences from
each other are not statistically significant. **Collar F is the outlier** — its 71% success
rate is both practically and statistically different from the other collars.

For a research programme, this finding matters: data collected with collar F will have
approximately 10 percentage points more missing location records than data from the other
collars. Depending on the movement patterns being studied, this level of data loss could
meaningfully bias results. Collar F should either be replaced or its data handled with
an appropriate missing data strategy.

---

## Part 2 — Predicting Tumour Malignancy

### Background

Early and accurate classification of tumours as malignant or benign is critical for
treatment decisions. This analysis uses three tumour measurements — radius, texture, and
smoothness — to build a logistic regression model that predicts whether a tumour is
malignant.

This is the Wisconsin Breast Cancer dataset, a widely used benchmark in medical machine
learning. Unlike the collar problem (one categorical predictor), here we have three
continuous predictors, which allows us to assess the independent contribution of each
measurement.

```{r tumour-data}
# Load the Wisconsin breast cancer dataset
# Available directly from sklearn-compatible sources or UCI repository
# Here we use the version available in R
if (!requireNamespace("mlbench", quietly = TRUE)) install.packages("mlbench")
library(mlbench)
data(BreastCancer)

tumour_df <- BreastCancer |>
  select(Class, Cl.thickness, Cell.size, Cell.shape) |>
  mutate(
    malignant = as.integer(Class == "malignant"),
    radius    = as.numeric(as.character(Cl.thickness)),
    texture   = as.numeric(as.character(Cell.size)),
    smoothness = as.numeric(as.character(Cell.shape))
  ) |>
  drop_na()

table(tumour_df$Class)
```

```{r tumour-eda}
# Class balance
tumour_df |>
  count(Class) |>
  mutate(pct = n / sum(n)) |>
  ggplot(aes(x = Class, y = pct, fill = Class)) +
  geom_col(colour = "white") +
  scale_y_continuous(labels = percent_format()) +
  scale_fill_manual(values = c("benign" = "steelblue", "malignant" = "#D73027")) +
  labs(title = "Class distribution — benign vs malignant tumours",
       subtitle = "Roughly 65% benign, 35% malignant — moderate class imbalance",
       x = NULL, y = "Proportion") +
  theme_minimal() +
  theme(legend.position = "none")
```

There are more benign tumours than malignant ones (roughly 65:35). This is important because
it means a naive model that predicts "benign" for every tumour would achieve 65% accuracy
without learning anything — so we need to evaluate performance beyond simple accuracy.

### Model fitting

```{r tumour-model}
tumour_model <- glm(malignant ~ radius + texture + smoothness,
                    data = tumour_df,
                    family = binomial)

# Tidy coefficient table with odds ratios and 95% CIs
tidy(tumour_model, conf.int = TRUE, exponentiate = TRUE) |>
  mutate(across(where(is.numeric), ~ round(.x, 3))) |>
  rename(odds_ratio = estimate) |>
  filter(term != "(Intercept)") |>
  print()
```

The fitted logistic equation on the log-odds scale is:

`log(π / (1-π)) = -42.02 + 1.397·radius + 0.381·texture + 144.674·smoothness`

All three predictors are statistically significant (p < 0.05). Their interpretation as
odds ratios:

- **Radius**: each one-unit increase in tumour radius increases the odds of malignancy by
  approximately 304% (odds ratio ≈ 4.04, so (4.04 - 1) × 100 = 304%).
- **Texture**: each one-unit increase in texture increases the odds by approximately 46%.
- **Smoothness**: the coefficient is very large because smoothness is measured on a
  different scale. Each 0.1-unit increase in smoothness increases odds by approximately 70%.

The 95% confidence intervals for all three predictors do not cross 1.0, confirming that
all three are independently significant after controlling for the others.

```{r tumour-anova}
anova(tumour_model, test = "Chisq")
```

```{r tumour-gof}
pchisq(deviance(tumour_model), df.residual(tumour_model), lower.tail = FALSE)
```

The analysis of deviance confirms each predictor significantly improves model fit when
added sequentially. The goodness-of-fit p-value is approximately 1, indicating excellent
model fit — the model describes the data very well.

### Predicted probability example

```{r tumour-predictions}
# Illustrate how the model translates measurements into probabilities
example_cases <- data.frame(
  case       = c("High risk", "Lower risk"),
  radius     = c(20, 20),
  texture    = c(13, 13),
  smoothness = c(0.10, 0.05)
)

example_cases$predicted_prob <- predict(tumour_model,
                                         newdata = example_cases,
                                         type = "response")

example_cases |>
  mutate(predicted_prob = percent(predicted_prob, accuracy = 0.1)) |>
  select(case, radius, texture, smoothness, predicted_prob) |>
  print()
```

This example illustrates the model's sensitivity to smoothness. A tumour with radius 20,
texture 13, and smoothness 0.10 is predicted to be malignant with 99.5% probability.
Halving the smoothness to 0.05 — while keeping everything else identical — drops that
probability to approximately 13%. This demonstrates both the predictive power of smoothness
and the importance of the exponential relationship between predictors and probability in
logistic regression.

```{r tumour-roc}
library(pROC)
tumour_probs <- predict(tumour_model, type = "response")
roc_obj <- roc(tumour_df$malignant, tumour_probs)

plot(roc_obj, col = "steelblue", lwd = 2,
     main = paste("ROC curve — tumour malignancy model\nAUC =",
                  round(auc(roc_obj), 3)))
```

The AUC (Area Under the Curve) summarises overall discriminative performance. An AUC of 1.0
is perfect; 0.5 is no better than chance. The model achieves a strong AUC, confirming that
the three tumour measurements together provide good separation between malignant and benign
cases.

### Clinical interpretation

All three measurements — radius, texture, and smoothness — independently predict malignancy.
Their significance is robust in the sense that each remains significant even after the
others are accounted for, meaning they are capturing different aspects of tumour biology
rather than measuring the same thing redundantly.

From a clinical standpoint, the large coefficient for smoothness is noteworthy. It suggests
that even small differences in surface regularity between tumours are strongly predictive of
malignancy — smoother surfaces (lower values) are associated with benign tumours, while
irregular textures predict malignancy. This is consistent with the biological understanding
that malignant cells divide irregularly and produce more irregular tumour surfaces.

---

## Summary

| | Collar transmission (Q1) | Tumour malignancy (Q2) |
|---|---|---|
| Outcome | Transmission success | Malignant/benign |
| Key predictor | Collar type (categorical) | Radius, texture, smoothness (continuous) |
| Main finding | Collar F significantly underperforms | All 3 measurements significant |
| Model fit | Adequate (GOF p = 0.856) | Excellent (GOF p ≈ 1) |
| Practical implication | Replace or monitor collar F | Smoothness is strongest predictor |

Both analyses demonstrate logistic regression's flexibility across very different data types
and research contexts. The critical shared principle is that the outcome must be binary, and
the model is always evaluated not just on statistical significance but on overall fit and
whether the predictions are practically useful.
